---
output: github_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(gridExtra)
library(forcats)

df <- read.table(file='mentorship.tsv', header = TRUE, sep = "\t")

# plot margins
MARGIN <- unit(c(0.6,0.3,1,0.3), "cm")

PCT <- function(D, d=NA){
  if(all(is.na(d)))
    return(sprintf("%.1f%%", nrow(D) * 100 / nrow(df)))
  else
    return(sprintf("%.1f%%", nrow(D) * 100 / nrow(d)))
}

# subtitle generator
GETSB <- function(D, X, SUBTITLE=NA){
  SB <- c()
  counter = 1

  for(variable in levels(factor(D[[X]]))){
    SB[counter] <- sprintf("%s: %d", variable, nrow(D[D[[X]] == variable, ]))
    counter = counter + 1
  }

  if(is.na(SUBTITLE))
    return(paste(SB, collapse=" / "))
  else
    return(paste(SUBTITLE, "\n\n(", paste(SB, collapse=" / "), ")", sep=""))
}

BOX <- function(D, X, Y, FILL=NA, FILL_LEGEND=NA, XTITLE=NA, YTITLE=NA, YL=NA, TITLE=NA, SUBTITLE=NA){
  p <- ggplot(data=D, aes(y=get(Y), x=get(X))) +
        {
          if(!is.na(FILL)) {
            geom_boxplot(aes(fill=get(FILL)), alpha=0.35, outlier.shape = NA, color="#00000022", width=0.25)
          }
        } +

        {
          if(!is.na(FILL_LEGEND)) {
            scale_fill_discrete(name = FILL_LEGEND)
          }
        } +

        {
          if(all(!is.na(YL))) {
             scale_y_continuous(labels=YL, breaks=1:length(YL))
          }
        } +

        geom_violin(fill="#505050", alpha=0.25, width=0.5, colour = '#505050') +
        geom_jitter(colour='#222222', width = 0.1, height = 0.25, size=1, alpha=0.5) +
        theme_bw() +
        ylab(YTITLE) +

        {
          if(is.na(FILL_LEGEND)) {
            theme(legend.position="none")
          } else {
            theme(legend.position="bottom")
          }
        } +

        {
          if(!is.na(XTITLE))
            xlab(XTITLE)
          else
            theme(axis.title.x=element_blank())
        } +

        {
          if(!is.na(YTITLE))
            ylab(YTITLE)
          else
            theme(axis.title.y=element_blank())
        } +

        labs(title = TITLE, subtitle = GETSB(D, X, SUBTITLE)) +
        theme(panel.border = element_blank(),
              plot.margin = MARGIN,
              axis.line.x = element_blank(),
              axis.text.x = element_text(face = "bold"),
              axis.text.y = element_text(face = "bold"),
              panel.grid.minor = element_line(colour = "#EFEFEF"),
              panel.grid.major.x = element_blank(),
              axis.line.y = element_blank(),
              axis.ticks.x = element_blank(),
              plot.title = element_text(color = "#222222", size = 14, face = "bold"),
              plot.subtitle = element_text(color = "#444444", size = 10))

  return(p)
}


BAR <- function(D, X, Y, XL=NA, FILL_LEGEND=NA, FILL_COLORS=NA, TITLE=NA, SUBTITLE=NA){
  p <- ggplot(data=D, aes(x=get(X), fill=get(Y))) +
        geom_bar(position="fill", stat="count") +
        theme_bw() +

        {
          if(!is.na(FILL_LEGEND) & all(!is.na(FILL_COLORS))) {
            scale_fill_manual(values=FILL_COLORS, name=FILL_LEGEND)
          } else if(is.na(FILL_LEGEND) & all(!is.na(FILL_COLORS))) {
            scale_fill_manual(values=FILL_COLORS)
          }
        } +

        {
          if(is.na(FILL_LEGEND)) {
            theme(legend.position="none")
          } else {
            theme(legend.position="bottom")
          }
        } +

        {
          if(all(!is.na(XL))) {
             scale_x_continuous(labels=XL, breaks=1:length(XL))
          }
        } +

        {
          if(is.na(TITLE))
            labs(subtitle = GETSB(D, X, SUBTITLE))
          else
            labs(title = TITLE, subtitle = GETSB(D, X, SUBTITLE))
        } +

        theme(panel.border = element_blank(),
              plot.margin = MARGIN,
              axis.title.y=element_blank(),
              axis.title.x=element_blank(),
              axis.line.x = element_blank(),
              axis.text.x = element_text(face = "bold"),
              axis.text.y = element_blank(),
              panel.grid.minor = element_blank(),
              panel.grid.major = element_blank(),
              axis.line.y = element_blank(),
              axis.ticks.x = element_blank(),
              axis.ticks.y = element_blank(),
              plot.title = element_text(color = "#222222", size = 14, face = "bold"),
              plot.subtitle = element_text(color = "#444444", size = 10))

  return(p)
}

BAX <- function(D, X, TITLE=NA, YTITLE=NA, XL=NA, SUBTITLE="", SKIPDATASUM=NA, ROTATEXL=NA, COLOR=NA){
  p <- ggplot(data=D, aes(x=get(X))) +
        geom_bar(position="identity", stat="count", fill=COLOR) +
        theme_bw() +

        {
        if(!is.na(YTITLE))
          ylab(YTITLE)
        else
          theme(axis.title.y=element_blank())
        } +

        {
          if(all(!is.na(XL))) {
             scale_x_continuous(labels=XL, breaks=1:length(XL))
          }
        } +

        {
        if(is.na(SKIPDATASUM))
          labs(title = TITLE, subtitle = GETSB(D, X, SUBTITLE))
        else
          labs(title = TITLE, subtitle = SUBTITLE)
        } +

        {
        if(is.na(ROTATEXL))
          theme(axis.text.x = element_text(face = "bold"))
        else
          theme(axis.text.x = element_text(face = "bold", angle = 90, hjust=1))
        } +

        theme(panel.border = element_blank(),
              plot.margin = MARGIN,
              axis.title.x=element_blank(),
              axis.line.x = element_blank(),
              panel.grid.minor = element_blank(),
              panel.grid.major.x = element_blank(),
              axis.line.y = element_blank(),
              axis.ticks.x = element_blank(),
              axis.ticks.y = element_blank(),
              plot.title = element_text(color = "#222222", size = 14, face = "bold"),
              plot.subtitle = element_text(color = "#444444", size = 10))

  return(p)
}

ARRANGE <- function(A, B, HEIGHTS=c(1,1)){
 gA <- ggplotGrob(A)
 gB <- ggplotGrob(B)
 maxWidth = grid::unit.pmax(gA$widths[2:5], gB$widths[2:5])
 gA$widths[2:5] <- as.list(maxWidth)
 gB$widths[2:5] <- as.list(maxWidth)
 return(grid.arrange(gA, gB, ncol=1, heights=HEIGHTS))
}

```

## Introduction to the survey data summary

!WARNING! [The survey is located here](https://forms.gle/RPksYdyihGuSBZMa8). **You still can take the survey or encourage others to do it**. Since the workflow that generates the figures and text below is [fully reproducible](https://github.com/merenlab/mentorship-survey-workflow), it is possible to update the results with additional data relatively quickly. If that ever happens, this note will reflect that with dates and significant changes in results if there are any. The data below cover a snapshot of the survey results. For your reference, the time stamp of the last answer in the survey at the time of the analysis was **`r tail(df, n=1)$timestamp`** (MM/DD/YYYY HH:MM:SS).

!NOTICE! This section offers a general summary of mentees and mentors.

The survey results analyzed below includes a total of **`r nrow(df)` responses** that reported a single experience between a mentor and an early career researcher (ECR). The definition of '**ECR**' in the survey includes PhD students, post-doctoral researchers, and assistant professors (i.e., those who typically are advised by others in academia and do not typically have 'job security'), and the text often uses the terms 'ECR' and 'mentee' interchangeably. In contrast, the term '**mentor**' defines someone who was officially responsible for the training of an ECR and include PhD or post-doc advisers, or more generally the principal investigator (PI), the group leader, or the director of a group affiliated with an academic institution. All data, including data for mentors (such as their career stage, gender, and minority status), are reported by their mentees.

### Disciplines

The vast majority of those who filled the survey were **early career researchers in sub-disciplines of Life Sciences**:

```{r discipline, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

D <- df
D$discipline <- factor(D$discipline, levels = c("Life Sciences", "Applied Sciences", "Physical Sciences", "Social Sciences", "Formal Sciences"))

BAX(D,
    X="discipline",
    TITLE="Science discipline",
    SUBTITLE="If you had to chose one, which one of the following categories would best describe your relationship\nwith science as an ECR or when you were an ECR?",
    COLOR="#66339999")
```

This is likely due to the limits of my reach on Twitter.

### Countries

```{r country_setup, echo=FALSE}
D <- df
D$country <- fct_infreq(D$country, ordered = NA)
all_countries <- GETSB(D, 'country')
```

Another limitation of the survey is the representation of countries. Although there were at least one person filled the survey from **a total of `r length(unique(df$country)) - 1` countries**, the number of individuals from each country was heavily skewed, and the vast majority of those who filled the survey worked in **the United States**:

```{r country, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

D <- df[df$country %in% names(table(df$country))[table(df$country) >=5], ]

D$country <- fct_infreq(D$country, ordered = NA)

BAX(D,
    X="country",
    SKIPDATASUM=TRUE,
    ROTATEXL=TRUE,
    TITLE="Countries represented five or more times in the survey",
    COLOR="#66336699")
```

The figure below displays the list of countries with five or more survey results, but for the sake of posterity, each country and the number of responses looked like this: `r all_countries`.


### Career stages

Most participants mentioned their experience as **graduate students**, and the large fraction of mentors mentees have reported on were **full professors**:

```{r mentee_mentor_career_stage, echo=FALSE, fig.width=9, fig.height=7, dpi=300}

D <- df
D$mentee_then <- factor(D$mentee_then, levels = c("Grad student", "Postdoc", "Assistant Prof"))

ARRANGE(
  BAX(D,
      X="mentee_then",
      TITLE="Mentee career stage",
      SUBTITLE="Reported by mentees",
      COLOR="#66339999"),

  BAX(D,
      X="mentor_then",
      TITLE="Mentor career stage",
      SUBTITLE="Reported by mentees",
      COLOR="#33669999"))

```


### Genders

A **larger fraction of the mentees were women**, and **a larger fraction of the mentors were men**.

```{r mentee_mentor_gender, echo=FALSE, fig.width=9, fig.height=7, dpi=300}

D <- df
genders <- c("Woman", "Man", "Queer / Non-conforming", "Prefer not to say")
D$mentee_gender <- factor(D$mentee_gender, levels = genders)
D$mentor_gender <- factor(D$mentor_gender, levels = genders)

ARRANGE(
    BAX(D,
        X="mentee_gender",
        TITLE="Mentee gender distribution",
        COLOR="#66339999"),

    BAX(D,
        X="mentor_gender",
        TITLE="Mentor gender distribution",
        SUBTITLE="(Reported by mentees)",
        COLOR="#33669999"),
    HEIGHTS = c(1,1))

```

While mentees who were men were shared equally between men and women mentors, a **larger proportion of women worked with mentors who were women**.

```{r mentor_mentee_gender, echo=FALSE, fig.width=9, fig.height=5, dpi=300}
D <- df[df$mentee_gender %in% c("Man", "Woman") & df$mentor_gender %in% c("Man", "Woman"), ]

BAR(D,
    X="mentor_gender",
    Y="mentee_gender",
    FILL_LEGEND = "Mentor gender",
    TITLE="Gender matching between mentees and mentors",
    FILL_COLORS = c("#143D59", "#F4B41A"))

```

!NOTICE! The survey question offered the following choices: 'Man', 'Woman', 'Genderqueer / Gender Non-Conforming', 'Questioning', 'Other', and 'Prefer not to say'. I used this set of choices based on the advice given [on this page](https://uxdesign.cc/designing-forms-for-gender-diversity-and-inclusion-d8194cf1f51) for *Limited Lists*. Another note worth highlighting is the fact that **sexual orientation was not a question in the survey**. The reason for that was the following: the more data one collects in a survey, the more likely it becomes that the combination of information can identify someone. As I planned to make the lightly processed raw data available along with a reproducible workflow, I decided to not collect any data on sexual orientation to maximize the anonymity of participants (even if I lost control of my online accounts) despite the fact that I believed that it may have been an important source of additional information.

### Minority status

```{r mentee_mentor_minority_data_prep, echo=FALSE}

D <- df

mentee_minority_pct = PCT(D[D$mentee_minority == "Yes", ], D[D$mentee_minority == "No", ])
mentor_minority_pct = PCT(D[D$mentor_minority == "Yes", ], D[D$mentor_minority == "No", ])

mentee_minority_mentor_minority = PCT(D[D$mentor_minority == "Yes" & D$mentee_minority == "Yes", ], D[D$mentor_minority == "Yes", ])
mentee_minority_mentor_majority = PCT(D[D$mentor_minority == "No" & D$mentee_minority == "Yes", ], D[D$mentor_minority == "No", ])

```

The survey also collected data on the minority status of mentees and mentors they worked with. The question was aimed to identify those who, except their gender, were a member of a minority group in their workplace based on ethnic background and/or religion. While **`r mentee_minority_pct` of mentees were a member of a minority group**, only **`r mentor_minority_pct` of mentors belonged to a minority group**, to the best of the knowledge of their mentees.

```{r mentee_mentor_minority, echo=FALSE, fig.width=9, fig.height=7, dpi=300}

ARRANGE(
  BAX(D,
      X="mentee_minority",
      TITLE="Mentee minority status",
      SUBTITLE="Except your gender, are you a member of a minority group in your workplace based on your\nethnic background and/or religion?",
      COLOR="#66339999"),
  BAX(D,
      X="mentor_minority",
      TITLE="Mentor minority status",
      SUBTITLE="Reported by mentees",
      COLOR="#33669999"))
```

The bars below show the matching of the minority status between mentees and mentors. While `r mentee_minority_mentor_minority` of the mentees who work with mentors that belong to a minority group also identify as minority, mentees who identify as minority make up only `r mentee_minority_mentor_majority` of the mentees who work with mentors who do not belong to a minority group.

```{r mentor_mentee_minoritry, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

BAR(D,
    X="mentor_minority",
    Y="mentee_minority",
    FILL_LEGEND = "Mentee was a member of a minority group?",
    FILL_COLORS = c("#143D59", "#F4B41A"),
    TITLE="Minority status matching between mentors and mentees",
    SUBTITLE="Except their gender, and to the best of your knowledge, was your mentor a member of a minority\ngroup in the workplace based on their ethnic background and/or religion?'")

```

Each column of the figure groups mentors based on their minority status (Yes/No answers on the x-axis are given to the question "was your mentor a member of a minority group"), and each color reports the minority status of the proportion of mentees who are working with these groups of mentors.

---

So far, the data show that the vast majority of people who filled the survey were life scientists from the United States, most of them were women, most of them reported their experience as graduate students, and most mentors were full professors during the mentor-mentee interaction. In addition we see that women are slightly more likely to be mentored by women, and those who identify as a member of a minority group are more likely to work with mentors they identify as a member of a minority group.

## Experiences of ECRs with their mentors

!NOTICE! This section offers a general summary of how ECRs felt about their previous experiences.

### The overall spectrum of experiences

Since one of the main goals of this survey was to capture the nature of experiences of early career researchers with their past or current academic mentors, the most critical question the survey asked to its participants was the following: **Considering its entirety and its influence on your well-being and career, how would you characterize your experience with your mentor?**

Here is how the responses looked like:

```{r experience_with_mentor_overall, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

D <- df

BAX(D,
    X="experience_with_mentor",
    TITLE="The overall experience of ECRs with their mentors",
    SUBTITLE="Considering its entirety and its influence on your wellbeing and career, how would you characterize\nyour experience with your mentor?",
    XL=c("Very Negative (1)", "Negative (2)", "Neutral (3)", "Positive (4)", "Very Positive (5)"),
    COLOR="#99446699")
```

This shows that `r PCT(df[df$experience_with_mentor > 3, ])` of all `r nrow(df)` participants reported a **positive experience** with their mentors, and **very negative experiences** were no more than `r PCT(df[df$experience_with_mentor == 1, ])` of the responses.

!NOTICE! OK. I will take a break here and tell you that I wrote the following part until the next subsection only after reading the experiences of mentees. I had not done it before to make sure they did not influence the direction of my preliminary analyses. After doing it, though, I had to come here and add the following text retrospectively.

Most scientists are aware that mentor-mentee relationships can occasionally go either extremes of the experience spectrum. And considering all the stories I personally have witnessed as an ECR either through social media or by watching the journey of other ECRs as mentees, I did not know what the distribution of experiences was going to look like. But my interpretation of these results changed as I continued to work on the article, especially after I started reading the words of mentees.

First, I found the large proportion of positive experiences pleasantly surprising. I even thought that it will likely serve as a relief also for you to learn that the vast majority of experiences among mentee-mentor pairs were positive.

Only later I realized that these numbers are not anywhere close to where we should set the bar as scientists. In fact, `r PCT(df[df$experience_with_mentor == 1, ])` is astonishingly large when one realizes the significance of very negative experiences on others' life and well-being. Here is an example that shook me:

<blockquote>
She was a covert narcissist who changed the goalposts to suit her narrative on a daily basis. Being in her lab made me suicidal. I can barely talk about this as it is still so traumatizing.
<div class="blockquote-author"><b>Postdoc</b> (♀) had a <span style="color:red;">very negative experience</span> in <b>United States</b><br />reporting on the <u>biggest shortcomings</u> of their mentor (♀)</div>
</blockquote>

As I went through the responses, my initial relief started to feel more and more embarrassing. The reason I decided to share this relatively off-topic personal experience of mine openly is because I believe this attitude is probably a significant part of the problem, and likely more common than those who did not suffer from the consequences of very negative experiences typically appreciate. It is much easier for those who are in the lucky majority to see the proportions of those who found themselves in the unlucky minority as mere statistics. `r PCT(df[df$experience_with_mentor == 1, ])` indeed makes a small fraction of all mentees in science. But people who are assaulted, or people who suffer from rare diseases make up a very small fraction of the society, too, and much less than `r PCT(df[df$experience_with_mentor == 1, ])` in most societies. But for those small number of people who are assaulted or diagnosed with a rare disease, life changes irreversibly.

This personal experience convinced me that this is the true power of this survey. The power of this survey is not the opportunities it enables for data analyses or summary of positive or negative experiences as a function of categories the survey fits people into. It is the words of those who lived these experiences, and the power and the magnitude of those words that can convince us to the importance of working towards an ideal where all mentee-mentor experiences are positive or very positive, and that there is a lot more to do to achieve that.

This section stil exists as I managed to fight off the desire to delete everything I've done until this realization, and let the text continue to try to offer a general summary of the data for posterity before leaving you with [the words of early career researchers](#the-words-of-early-career-researchers).

### Commonality of experiences within a group

In the most general sense, ECRs believe that their experiences with their mentors were common to all mentees trained by their mentor:

```{r experience_with_mentor_common_in_group_01, echo=FALSE, fig.width=9, fig.height=5, dpi=300}
D <- df

BAX(D,
    X="mentee_experience_was_common",
    TITLE="The commonality of mentee experiences in a group",
    SUBTITLE="Was your experience with your mentor comparable to the experience of the other trainees\nwho worked with them?",
    XL=c("Very specific (1)", "Somewhat specific (2)", "Neutral (3)", "Somewhat common (4)", "Very common (5)"),
    COLOR="#99446699")

v_pos_specific <- PCT(D[D$mentee_experience_was_common_summary == "Specific to few" & D$experience_with_mentor == "5", ], D[D$experience_with_mentor == "5" & D$mentee_experience_was_common_summary != "", ])
v_neg_specific  <- PCT(D[D$mentee_experience_was_common_summary == "Specific to few" & D$experience_with_mentor == "1", ], D[D$experience_with_mentor == "1" & D$mentee_experience_was_common_summary != "", ])
```

But interestingly, the extent of commonness of experiences were not uniform across distinct categories of mentorship outcomes. While only `r v_pos_specific` of the mentees who had a very positive experience with their mentors thought that their experience was specific to a few, as much as `r v_neg_specific` of the mentees who had a very negative experience with their mentor believed that their experience was specific to few people trained by the same mentor:

```{r experience_with_mentor_common_in_group_02, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

D <- df[df$mentee_experience_was_common_summary != "", ]

BAR(D,
    X="experience_with_mentor",
    Y="mentee_experience_was_common_summary",
    FILL_LEGEND = "",
    FILL_COLORS = c("#FFC107", "#DD0202"),
    TITLE="The commonality of mentee experiences in a group",
    XL=c("Very Negative (1)", "Negative (2)", "Neutral (3)", "Positive (4)", "Very Positive (5)"),
    SUBTITLE="Was your experience with your mentor comparable to the experience of the other trainees\nwho worked with them?")
```

I find this to be an important result.

The following is one way to interpret these data: mentors who are able to provide positive experiences to their mentees do it consistently, as most of the mentees who reported very positive experiences also thought that their experience was common to most remembers of the group. Even if the data in this survey is powered enough to make a statement on this, there may be multiple perceptional biases here. For instance, those who are having a very positive experience with their mentors may be unable to appreciate what some of the other mentees may be going through. In contrast, it may be easier for those who are suffering with their mentors to recognize others are not suffering from the same problems.

Nevertheless, the marked increase in the proportion of mentees who report very negative and yet specific experiences suggests that even mentors who generally are able to have positive experiences with some of their mentees can have very negative experiences. Putting this into words almost sounds like a tautology to me since it should not surprise anyone that a single mentorship strategy will unlikely work for everyone equally well. Going through mentee experiences later, I could find evidence for this obvious point in survey results itself where while some mentees praised a behavior, others saw it as a shortcoming. For instance, the following is reported by a post-doc who believed the very hands-off approach was the biggest *shortcoming* of their mentor's mentorship style:

<blockquote>
He was very hands off. He would not insist about deadlines for example. This made me miss some deadlines for fellowships etc. because he did not reminded me or helped me made the deadline.
<div class="blockquote-author"><b>Postdoc</b> (♂), <b>United States</b><br />reporting on the <u>biggest shortcomings</u> of their mentor (♂)</div>
</blockquote>

In contrast, this one is reported by a graduate student who believes the same strategy, albeit challenging, was the most admirable trait of their mentor:

<blockquote>
Completely hands-off approach, which was really difficult at first, but in the end I believe made me a more self-sufficient researcher and ready to establish myself in academia
<div class="blockquote-author"><b>Grad student</b> (♂), <b>United States</b><br />reporting on the <u>most admirable qualities</u> of their mentor (♂)</div>
</blockquote>

I think this foreshadows that some of the negative experiences may be due to the mismatch between what the mentee is expected from their mentor, and what the mentor is able to offer to their mentees. ONLY IF THERE WAS A WAY for mentors to be upfront about their mentorship strategy, so mentees could have an opportunity to find better matches.

We will get back to this at the end of this section.

### The impact of the mentor to the next career stage

```{r experience_with_mentor_future_prep, echo=FALSE}
D <- df

levels(D$mentee_then) <- levels(factor(D$mentee_current))

D <- D[D$mentee_current != D$mentee_then, ]
```

Of all `r nrow(df)` ECRs who took the survey, `r nrow(D)` had a major change in their career stage since the experience they reported with their mentors: graduate students became post-docs, post-docs became professors, or they left the academic life behind and moved on to other endeavors.

The subset of individuals who had a change in their career status offers an opportunity to investigate whether the data suggest anything regarding what did ECRs do next as a function of their experience with their mentors. One of the key insights one could gain here is whether or not those who had very negative experiences with their mentors were more likely to end up leaving the academic life behind.

According to the survey data, there was a very slight increase in the proportion of individuals who had very negative experiences with their mentors and left academia:

```{r experience_with_mentor_future, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

D <- df
levels(D$mentee_then) <- levels(factor(D$mentee_current))

D <- D[D$mentee_current != D$mentee_then, ]

BAR(D,
    X="experience_with_mentor",
    Y="mentee_current_summary",
    FILL_LEGEND = "The current career stage:",
    FILL_COLORS = c("#0f9257", "#F7B32B", "#577AA3"),
    TITLE="The influence of the experience with mentors to career trajectories",
    XL=c("Very Negative (1)", "Negative (2)", "Neutral (3)", "Positive (4)", "Very Positive (5)"),
    SUBTITLE="What did ECRs do next once their experience with their mentor was over?")
```

Even after *negative* or *very negative* experiences with a mentor, mentees seem to have moved on with their careers in academia in comparable proportions to those who had positive or very positive experiences. But these results on careers are to be taken with *two* grains of salt as this survey may be grossly underestimating the number of those who may have left the academic life (since they are less likely to see the survey and/or have the energy to fill it).

## Factors that are associated with positive vs negative outcomes

!NOTICE! This section aims to investigate strong associations by benchmarking various categorical data and positive or negative outcomes of mentor-mentee relationships.

### Gender of the mentee or the mentor

Does gender plays a role in the outcomes of mentee-mentor pairings? Does our collective mind have room for yet another [blatantly incorrect](https://www.nature.com/articles/s41467-020-19723-8) interpretation of the matter???

Well, thankfully our survey suggests that the overall experiences of mentees with their mentors do not seem to correlate with the gender of the mentee:

```{r mentee_gender_mentor_exp, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

D <- df[df$mentee_gender %in% c("Man", "Woman") & df$mentor_gender %in% c("Man", "Woman"), ]

# wilcox.test(D[D$mentee_gender == "Man", ]$experience_with_mentor, D[D$mentee_gender == "Woman", ]$experience_with_mentor)

BOX(D,
    X="mentee_gender",
    Y="experience_with_mentor",
    FILL="mentor_gender",
    FILL_LEGEND = "Mentor gender",
    YTITLE=NA,
    XTITLE=NA,
    YL=c("Very Negative", "Negative", "Neutral", "Positive", "Very Positive"),
    TITLE="Mentee Gender vs Experience With Mentor",
    SUBTITLE="Considering its entirety and its influence on your wellbeing and career, how would you characterize\nyour experience with your mentor?")

```
The distribution of positive and negative experiences seem to be quite uniform across genders (just to note, while the x-axis shows the gender of mentees, the inline box plots highlight the distribution of the gender of mentors within a given category).

One could also ask whether the overall experiences of mentees with their mentors correlate with the gender of their mentor:

```{r mentor_gender_mentor_exp, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

D <- df[df$mentee_gender %in% c("Man", "Woman") & df$mentor_gender %in% c("Man", "Woman"), ]

# wilcox.test(D[D$mentor_gender == "Man", ]$experience_with_mentor, D[D$mentor_gender == "Woman", ]$experience_with_mentor)

BOX(D,
    X="mentor_gender",
    Y="experience_with_mentor",
    FILL="mentee_gender",
    FILL_LEGEND = "Mentee gender",
    XTITLE=NA,
    YTITLE=NA,
    YL=c("Very Negative", "Negative", "Neutral", "Positive", "Very Positive"),
    TITLE="Mentor Gender vs Experience With Mentor",
    SUBTITLE="To see if mentor gender has a general effect")
```
As somewhat foreshadowed by the previous data, there does not seem to be any correlation here, either.

Gender is out of the way. Moving on.

### Mentor group size

Since the increasing number of mentees will decrease the availability of a mentor to each of their trainees, group size (as in the number of trainees a mentor is responsible) is a reasonable factor to consider as a function of mentee experience.

And it does seem that group size and overall mentee experience have a weak, but interesting relationship:

```{r mentor_group_size, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

D <- df

D$mentor_num_trainees <- factor(D$mentor_num_trainees, levels = c("<5", "5-10", "10-15", ">15"))

BOX(D,
    X="mentor_num_trainees",
    Y="experience_with_mentor",
    FILL="mentor_num_trainees",
    FILL_LEGEND = NA,
    YTITLE=NA,
    XTITLE=NA,
    YL=c("Very Negative", "Negative", "Neutral", "Positive", "Very Positive"),
    TITLE="Number of trainees vs experience with mentor",
    SUBTITLE="How many trainees your mentor was responsible for during the time\nyou worked with them?")

```

It seems the experiences of mentees in groups of less than five people are relatively more uniformly distributed across different outcome categories. While mentors who are responsible for more than 15 people offer very few very positive or very negative mentorship experiences. And the sweet spot seems to be those who work with 5 to 10 people.

### Awareness of mentors on their strengths and shortcomings

A couple of questions in the survey asked mentees whether they thought their mentors were aware of their strengths and shortcomings in mentorship. According to mentees, the extent of awareness among their mentors were not uniformly distributed across their strengths or shortcomings:

```{r mentor_awareness, echo=FALSE, fig.width=9, fig.height=7, dpi=300}
awareness <- c("Aware", "Not Aware")

D <- df[df$mentor_awareness_on_strengths_summary %in% awareness,  ]
R <- df[df$mentor_awareness_on_shortcomings_summary %in% awareness,  ]

D$mentor_awareness_on_strengths_summary <- factor(D$mentor_awareness_on_strengths_summary, levels = awareness)
R$mentor_awareness_on_shortcomings_summary <- factor(R$mentor_awareness_on_shortcomings_summary, levels = awareness)

grid.arrange(
    BAX(D,
        X="mentor_awareness_on_strengths_summary",
        TITLE="Strengths in mentorship",
        SUBTITLE = "Do you think your mentor was aware of their strengths in mentoring trainees?",
        COLOR="#66339999"),

    BAX(R,
        X="mentor_awareness_on_shortcomings_summary",
        TITLE="Shortcomings in mentorship",
        SUBTITLE = "Do you think your mentor was aware of their shortcomings in mentoring trainees?",
        COLOR="#33669999"))

```

In the eyes of mentees, mentors were much less aware of their shortcomings than their strengths in mentorship. But there was a marked difference between the overall experience of mentees based on the extent of awareness of mentors in their skills:

```{r mentee_awareness, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

D <- df[df$mentor_gender %in% c("Man", "Woman") & df$mentor_awareness_on_shortcomings_summary %in% c("Aware", "Not Aware"), ]

BOX(D,
    X="mentor_awareness_on_shortcomings_summary",
    Y="experience_with_mentor",
    FILL="mentor_gender",
    FILL_LEGEND = "Mentor gender",
    YTITLE=NA,
    XTITLE=NA,
    YL=c("Very Negative", "Negative", "Neutral", "Positive", "Very Positive"),
    TITLE="Mentor awareness on shortcomings in mentorship",
    SUBTITLE="Do you think your mentor was aware of their shortcomings in mentoring trainees")

```

Mentors who were aware of their shortcomings in mentoring had more likely to have mentees with whom they had very positive relationships.

Here is a data-driven tip: If you are an ECR considering to start working with a new mentor, ask them what they think is their biggest shortcoming in mentorship. Then ask the same question to their trainees. It may be a good sign if they match. If you are a mentor, think about what might be your shortcomings in mentoring people.

### Mentors asking for feedback

If mentors who are aware of their shortcomings in mentoring in fact offer a better mentorship experience to their mentees, then, as mentors, we do have quite a simple solution to recover from our lack of awareness of our shortcomings by doing this one simple thing: asking those who we mentor for feedback.

But according to the survey results, most mentors have never asked for feedback on their mentorship style:

```{r mentor_asking_feedback_01, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

D <- df
D$mentor_asking_feedback <- factor(D$mentor_asking_feedback, levels = c("Frequently", "Every six months", "Every year", "Once every few years", "Never"))

BAX(D,
      X="mentor_asking_feedback",
      TITLE="Mentors asking for feedback on their mentorship style",
      SUBTITLE="How frequently mentors ask for feedback from their trainees on their mentorship style?\n",
      COLOR="#66339999")

never_asked_feedback = PCT(D[D$mentor_asking_feedback == "Never", ])
```

In fact, as much as `r never_asked_feedback` of mentees have worked with mentors who never even once asked for any feedback on their mentorship style. This percentage is much higher than the percentage of negative experiences, which means that not all mentors who are not asking for feedback on their mentorship style are failing to provide positive experiences. But the data are very clear on the fact that almost the entirety of very negative mentee experiences seem to be associated with mentors who never ask for feedback:

```{r mentor_asking_feedback_03, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

BOX(D,
    X="mentor_asking_feedback",
    Y="experience_with_mentor",
    FILL= "mentor_asking_feedback",
    FILL_LEGEND = NA,
    YTITLE=NA,
    XTITLE=NA,
    YL=c("Very Negative", "Negative", "Neutral", "Positive", "Very Positive"),
    TITLE="Mentee experience with mentors who ask for feedback",
    SUBTITLE="What is the correlation between mentee experience with mentors requesting feedback?\n")

```

It is indeed extremely difficult to establish a causative relationship here. But when it comes to asking for feedback, it seems women do it slightly more than men, Assistant Professors do it slightly more than Professors, and mentors who are considered to be 'good' by their colleagues (according to mentees) do it more than those who are not particularly considered good by their colleagues:

```{r mentor_x_asking_feedback, echo=FALSE, fig.width=10, fig.height=10, dpi=300}
Countries <- c("United States", "Germany", "United Kingdom")
D <- df
R <- df[df$country %in% Countries, ]
R$country <- factor(R$country, levels = Countries)

grid.arrange(
  BAR(df[df$mentor_gender %in% c("Man", "Woman"), ],
      X="mentor_gender",
      Y="mentor_has_ever_asked_feedback",
      TITLE="Mentor gender",
      FILL_COLORS = c("#474341", "#FFC107")),

  BAR(D,
      X="mentor_minority",
      Y="mentor_has_ever_asked_feedback",
      TITLE="Mentor minority status",
      FILL_COLORS = c("#474341", "#FFC107")),

  BAR(D[D$mentor_seen_by_colleagues != "No comment", ],
      X="mentor_seen_by_colleagues",
      Y="mentor_has_ever_asked_feedback",
      TITLE="Mentor seen by colleagues",
      FILL_COLORS = c("#474341", "#FFC107")),

  BAR(D,
      X="code_of_conduct",
      Y="mentor_has_ever_asked_feedback",
      TITLE="Mentor has code of conduct",
      FILL_COLORS = c("#474341", "#FFC107")),

  BAR(D,
      X="mentor_then",
      Y="mentor_has_ever_asked_feedback",
      FILL_COLORS = c("#474341", "#FFC107"),
      TITLE="Mentor career stage",
      SUBTITLE=NA,
      FILL_LEGEND = "Mentor asked feedback at least once"),

  BAR(R,
      X="country",
      Y="mentor_has_ever_asked_feedback",
      TITLE="Mentor country of work",
      FILL_COLORS = c("#474341", "#FFC107"),
      FILL_LEGEND = "Mentor asked feedback at least once")
  )
```

### Are mentors making it easy to raise mentorship concerns

If a mentor is not explicitly asking for feedback from their mentees on the mentorship they are receiving, it may not be the end of the world, if they are at least making it easier for their mentees to raise their concerns. However, most mentors seem to make it difficult for their mentees to raise their concerns with them:

```{r mentor_mentorship_concerns_01, echo=FALSE, fig.width=9, fig.height=5, dpi=300}
easiness <- c( "Made it easy", "Made it difficult")

J <- df[df$mentor_makes_easy_mentees_to_raise_concerns_summary %in% easiness,  ]

J$mentor_makes_easy_mentees_to_raise_concerns_summary <- factor(J$mentor_makes_easy_mentees_to_raise_concerns_summary, levels = easiness)

BAX(J,
    X="mentor_makes_easy_mentees_to_raise_concerns_summary",
    TITLE="Making it easy to raise concerns",
    SUBTITLE = "Do you think your mentor made it easy for their trainees to raise their concerns\nabout the mentorship they were receiving?",
    COLOR="#66339999")
```

And the association between positive mentee experiences and difficulty in raising concerns with mentorship they receive is quite stark:

```{r mentor_mentorship_concerns_02, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

BOX(J,
    X="mentor_makes_easy_mentees_to_raise_concerns_summary",
    Y="experience_with_mentor",
    FILL= "mentor_makes_easy_mentees_to_raise_concerns_summary",
    FILL_LEGEND = "",
    YTITLE=NA,
    XTITLE=NA,
    YL=c("Very Negative", "Negative", "Neutral", "Positive", "Very Positive"),
    TITLE="Mentee experience with mentors who makes it easy to raise concerns",
    SUBTITLE="Do you think your mentor made it easy for their trainees to raise their concerns\nabout the mentorship they were receiving?")

```

Almost none of the mentees had a negative experience with a mentors who made it easier to raise concerns. Creating an environment where it doesn't take much to be able to raise concerns is indeed a way to ensure negative experiences do not evolve to very negative ones.

```{r mentor_mentorship_concerns_03, echo=FALSE, fig.width=10, fig.height=10, dpi=300}
Countries <- c("United States", "Germany", "United Kingdom")
D <- J
R <- J[J$country %in% Countries, ]
R$country <- factor(R$country, levels = Countries)

grid.arrange(
  BAR(D[D$mentor_gender %in% c("Man", "Woman"), ],
      X="mentor_gender",
      Y="mentor_makes_easy_mentees_to_raise_concerns_summary",
      TITLE="Mentor gender",
      FILL_COLORS = c("#474341", "#FFC107")),

  BAR(D,
      X="mentor_minority",
      Y="mentor_makes_easy_mentees_to_raise_concerns_summary",
      TITLE="Mentor minority status",
      FILL_COLORS = c("#474341", "#FFC107")),

  BAR(D[D$mentor_seen_by_colleagues != "No comment", ],
      X="mentor_seen_by_colleagues",
      Y="mentor_makes_easy_mentees_to_raise_concerns_summary",
      TITLE="Mentor seen by colleagues",
      FILL_COLORS = c("#474341", "#FFC107")),

  BAR(D,
      X="code_of_conduct",
      Y="mentor_makes_easy_mentees_to_raise_concerns_summary",
      TITLE="Mentor has code of conduct",
      FILL_COLORS = c("#474341", "#FFC107")),

  BAR(D,
      X="mentor_then",
      Y="mentor_makes_easy_mentees_to_raise_concerns_summary",
      FILL_COLORS = c("#474341", "#FFC107"),
      TITLE="Mentor career stage",
      SUBTITLE=NA,
      FILL_LEGEND = "Mentor encouraging feedback"),

  BAR(R,
      X="country",
      Y="mentor_makes_easy_mentees_to_raise_concerns_summary",
      TITLE="Mentor country of work",
      FILL_COLORS = c("#474341", "#FFC107"),
      FILL_LEGEND = "Mentor encouraging feedback")
  )
```

But it seems no category of mentors seem to do it better than others. Although, those who have a code of conduct seem to be particularly successful at making their mentees feel that they can easily raise their concerns.

Which brings us to the final point here.

### Mentor has a lab culture or code of conduct document

I think it is safe to assume that most negative experiences of mentees took place despite the fact that the vast majority of mentors and mentees are doing their best. But our best is clearly not enough.

Data in this survey suggest that the match between mentors and mentees play a key role in defining the outcomes of mentee-mentor relationships. Fortunately there is a way to ensure that match early on by creating a *code of conduct* or a *lab culture & expectations* document to share with perspective mentees. These documents help mentors to be upfront about their mentorship strategy and their key expectations from the members of their group, so mentees could have an opportunity to find better matches.

As of today, such documents seem to be very uncommon among mentors:

```{r mentor_code_of_conduct_01, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

D <- df
D$code_of_conduct <- factor(D$code_of_conduct, levels = c("Yes", "No"))

BAX(D,
    X="code_of_conduct",
    TITLE="Mentor gives an idea about their style through a 'code of conduct'",
    SUBTITLE="Did your mentor provide a public document defining their code of conduct, expectations,\nand/or the group culture?",
    COLOR="#66339999")
```

Yet, the experience of trainees with mentors that provided a code of conduct seems to be much more positive:

```{r mentor_code_of_conduct_02, echo=FALSE, fig.width=9, fig.height=5, dpi=300}

D <- df

# wilcox.test(D[D$code_of_conduct == "Yes", ]$experience_with_mentor,
#             D[D$code_of_conduct == "No", ]$experience_with_mentor)

D$code_of_conduct <- as.character(D$code_of_conduct)
D$code_of_conduct <- factor(D$code_of_conduct, levels = c("Yes", "No"))

BOX(D,
    X="code_of_conduct",
    Y="experience_with_mentor",
    FILL="code_of_conduct",
    FILL_LEGEND = NA,
    YTITLE=NA,
    XTITLE=NA,
    YL=c("Very Negative", "Negative", "Neutral", "Positive", "Very Positive"),
    TITLE="Mentee experience with mentors who provide a code of conduct",
    SUBTITLE="Did your mentor provide a public document defining their code of conduct, expectations,\nand/or the group culture?")

```

And finally, the likelihood of offering code of conduct does not seem to be associated with pretty much anything, even though there is a slight increase among assistant professors from the US:

```{r mentor_x_code_of_conduct, echo=FALSE, fig.width=10, fig.height=10, dpi=300}
Countries <- c("United States", "Germany", "United Kingdom")
D <- df
R <- df[df$country %in% Countries, ]
R$country <- factor(R$country, levels = Countries)

grid.arrange(
  BAR(df[df$mentor_gender %in% c("Man", "Woman"), ],
      X="mentor_gender",
      Y="code_of_conduct",
      TITLE="Mentor gender",
      FILL_COLORS = c("#474341", "#FFC107")),

  BAR(D,
      X="mentor_minority",
      Y="code_of_conduct",
      TITLE="Mentor minority status",
      FILL_COLORS = c("#474341", "#FFC107")),

  BAR(D[D$mentor_seen_by_colleagues != "No comment", ],
      X="mentor_seen_by_colleagues",
      Y="code_of_conduct",
      TITLE="Mentor seen by colleagues",
      FILL_COLORS = c("#474341", "#FFC107")),

  BAR(D,
      X="mentor_has_ever_asked_feedback",
      Y="code_of_conduct",
      TITLE="Mentor has ever asked feedback",
      FILL_COLORS = c("#474341", "#FFC107")),

  BAR(D,
      X="mentor_then",
      Y="code_of_conduct",
      FILL_COLORS = c("#474341", "#FFC107"),
      TITLE="Mentor career stage",
      SUBTITLE=NA,
      FILL_LEGEND = "Mentor shares code of conduct"),

  BAR(R,
      X="country",
      Y="code_of_conduct",
      TITLE="Mentor country of work",
      FILL_COLORS = c("#474341", "#FFC107"),
      FILL_LEGEND = "Mentor shares code of conduct")

  )
```

I believe it is most critical for mentors to write these documents without reading documents written by others. Because I believe the most important role of these documents is that they push mentors to actually think about who they think they are (or perhaps more accurately, who they wish to be as mentors) and what they want their group to radiate as a cultural and intellectual gathering. So it is an extremely useful exercise for anyone who with to challenge themselves regardless of their career stage.

While I think mentors should draft their documents without reading others, here is a list of such documents I collected online that belong to various groups if you need some inspiration:

* [Lab Culture and Expectations](https://merenlab.org/culture/), [Meren Lab](http://merenlab.org), University of Chicago.
* [Lab Expectations](https://drkatlab.wordpress.com/resources/lab-expectations/), [Milligan-Myhre Lab](https://drkatlab.wordpress.com/), University of Connecticut.
* [Values and philosophy](https://lefo.ca/vie-de-labo/?lang=en), [Laboratoire d’écologie fonctionnelle végétale](https://lefo.ca/?lang=en), l’Université de Montréal.
* [Lab Policies and Procedures](https://cpb-us-e2.wpmucdn.com/faculty.sites.uci.edu/dist/4/340/files/2021/02/Lab_Policies_Draft_20210125.pdf), [Marine Biodiversity Lab](https://faculty.sites.uci.edu/biodiversity/), UC Irvine.
* [Lab Handbook](https://cobeylab.github.io/lab_handbook/), [Cobey Lab](https://cobeylab.uchicago.edu/), University of Chicago.
* [Lab Policies and Tips](http://www.avasthilab.org/lab-policies-and-tips/), [Avasthi Lab](http://www.avasthilab.org/), Dartmouth College.
* [Lab philosophy](https://maestrelab.com/en/lab-philosophy/), [Dryland Ecology
and Global Change Lab](https://maestrelab.com/en/), University of Alicante.
* [Lab Manual & Policies](http://big-data-biology.org/manual/), [Big Data Biology Lab](http://big-data-biology.org/), Fudan University.
* [Trainee Orientation Document](https://github.com/Gibbons-Lab/manuals-and-policies/blob/master/README.md), [Microbial Systems Ecology and Evolution Lab](https://gibbons.isbscience.org/), Institute for Systems Biology.

If you are a mentor, please consider writing a write a *code of conduct* or a *lab culture & expectations* document, and put it on your web page. I promise you that it will be an exercise that is both fun, and useful.
